### 2025-11-20 YEAR2/PHY293_LectureNotes/main.tex
LTNS

    \begin{equation}
        \int_{-\infty}^{\infty} |\psi(x,t)|^2 \, dx = 1
    \end{equation}

@@

\end{document}

>>>>

    \begin{equation}
        \int_{-\infty}^{\infty} |\psi(x,t)|^2 \, dx = 1
    \end{equation}
    , where $|\psi(x,t)|^2$ is:
    $$
        |\psi(x,t)|^2 = \psi^*(x,t) \psi(x,t)
    $$
    Here, \( \psi^*(x,t) \) is the complex conjugate of

@@

\section{The Schrödinger Equation in One-Dimensional Space}
\paragraph{Standing Waves and Stationary States}
\begin{itemize}
    \item A particle confined in space behaves like a standing wave.
    \item For a stationary state:
    \[
        \Psi(x,t) = \psi(x)e^{-iEt/\hbar},
    \]
    giving a time-independent probability density $|\Psi|^2 = |\psi|^2$.
    \item Superpositions of stationary states produce time-dependent densities.
\end{itemize}
\begin{definition}[Time Dependent Schrödinger Equation]
    The time-dependent Schrödinger equation describes how the quantum state of a physical system changes over time. It is given by:
    \begin{equation}
        i \hbar \frac{\partial \Psi(x,t)}{\partial t} = -\frac{\hbar^2}{2m} \frac{\partial^2 \Psi(x,t)}{\partial x^2} + V(x) \Psi(x,t)
    \end{equation}
    where \( \Psi(x,t) \) is the wave function of the particle, \( m \) is the mass of the particle, \( V(x) \) is the potential energy as a function of position, and \( \hbar \) is the reduced Planck's constant.

\end{definition}
\begin{example}[Particle in a Box]
    Consider a particle confined in a one-dimensional box of length \( L \) with infinitely high potential walls at \( x = 0 \) and \( x = L \). The potential \( V(x) \) is given by:
    $$
        V(x) = \begin{cases}
            0 & \text{for } 0 < x < L \\
            \infty & \text{otherwise}
        \end{cases}
    $$
    The time-independent Schrödinger equation inside the box is:
    $$
        -\frac{\hbar^2}{2m} \frac{d^2 \psi(x)}{dx^2} = E \psi(x)
    $$
    The general solution to this equation is:
    $$
        \psi(x) = A \sin(kx) + B \cos(kx)
    $$
    where \( k = \sqrt{\frac{2mE}{\hbar^2}} \).

    Applying the boundary conditions \( \psi(0) = 0 \) and \( \psi(L) = 0 \), we find that \( B = 0 \) and \( kL = n\pi \) for \( n = 1, 2, 3, \ldots \). Thus, the allowed wave numbers are:
    $$
        k_n = \frac{n\pi}{L}
    $$
    The corresponding energy levels are quantized:
    $$
        E_n = \frac{\hbar^2 k_n^2}{2m} = \frac{n^2 \pi^2 \hbar^2}{2mL^2}, \quad n = 1, 2, 3, \ldots
    $$
    The normalized wave functions are:
    $$
        \psi_n(x) = \sqrt{\frac{2}{L}} \sin\left(\frac{n\pi x}{L}\right)
    $$
    Therefore, the stationary states of the particle in the box are given by:
    $$        
    \Psi_n(x,t) = \sqrt{\frac{2}{L}} \sin\left(\frac{n\pi x}{L}\right) e^{-iE_n t/\hbar}
    $$
\end{example}

\begin{definition}[Energy Quantization]
    \begin{itemize}
    \item Only discrete energies $E_n$ yield normalizable solutions satisfying boundary conditions.
    \item Examples:
    \[
        \begin{aligned}
        U(x) &= 0 &\text{(infinite square well)}\\
        U(x) &= \tfrac{1}{2}kx^2 &\text{(harmonic oscillator)}\\
        U(r) &= -\frac{ke^2}{r} &\text{(hydrogen atom)}
        \end{aligned}
    \]
\end{itemize}

\end{definition}
\subsection{Eigenstates and Observables}
\begin{definition}[Ovservables and Operators]
    In quantum mechanics, physical observables such as position, momentum, and energy are represented by operators that act on the wave function. An operator is a mathematical entity that transforms one function into another. For example:
    \begin{itemize}
        \item Position operator: \( \hat{x} = x \)
        \item Momentum operator: \( \hat{p} = -i\hbar \frac{d}{dx} \)
        \item Hamiltonian (energy) operator: \( \hat{H} = -\frac{\hbar^2}{2m} \frac{d^2}{dx^2} + V(x) \)
    \end{itemize}
\end{definition}
\paragraph{From Schrödinger Equation to Observables}
The time-independent Schrödinger equation (TISE) governs quantum systems:
\begin{equation}
-\frac{\hbar^2}{2m}\frac{d^2\psi(x)}{dx^2} + U(x)\psi(x) = E\psi(x)
\end{equation}
Notice, that it is an \textbf{eigenvalue equation}.

\paragraph{Eigenvalue Equation}
When an operator $\hat{A}$ acts on a special function called an \textbf{eigenfunction} $\psi(x)$, it returns the function scaled by a measurable number called an \textbf{eigenvalue} $\lambda$:
\begin{equation}
\hat{A}\psi(x) = \lambda\psi(x)
\end{equation}

The TISE is precisely this structure, where $\hat{H}$ (the Hamiltonian) is the operator and $E$ is the eigenvalue:
\begin{equation}
    \hat{H}\psi(x) = E\psi(x)
\end{equation}

\paragraph{Common Operators} Below is a table of common quantum operators:
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
\textbf{Observable} & \textbf{Operator} & \textbf{Eigenvalue} \\
\hline
Position & $\hat{x} = x$ & $x_0$ \\
Momentum & $\hat{p} = -i\hbar\frac{d}{dx}$ & $p_0$ \\
Potential Energy & $\hat{V} = V(x)$ & $V(x_0)$ \\
Kinetic Energy & $\hat{K} = -\frac{\hbar^2}{2m}\frac{d^2}{dx^2}$ & $K$ \\
Energy (Hamiltonian) & $\hat{H} = -\frac{\hbar^2}{2m}\frac{d^2}{dx^2} + V(x)$ & $E$ \\
\hline
\end{tabular}
\end{center}
The choice of $V(x)$ depends on the physical system being modeled:
\begin{itemize}
    \item Infinite square well: $V(x) = 0$ inside the well, $V(x) = \infty$ outside.
    \item 
    \item Harmonic oscillator: $V(x) = \frac{1}{2}kx^2$.
    \item Hydrogen atom: $V(r) = -\frac{ke^2}{r}$.
\end{itemize}
\begin{definition}[Expected Value]
In quantum mechanics, the expectation value of an observable represented by operator $\hat{A}$ is:
\begin{equation}
    \langle A \rangle = \int_{-\infty}^{\infty} \psi^*(x) \, \hat{A}\psi(x) \, dx
\end{equation}
where $\psi(x)$ is the wave function of the system, and $\psi^*(x)$ is its complex conjugate. The expectation value represents the average outcome of measurements of the observable $A$ on a large number of identically prepared systems.
\end{definition}
\begin{example}
    A particle of mass \( m \) is confined in a one-dimensional infinite potential well of width \( L \). Its wave function is given by:
    $$
        \psi_n(x) = Ae^{x/L} \quad \text{for } 0 < x < L
    $$
    and \( \psi_n(x) = 0 \) elsewhere. Determine $\langle x \rangle$ for the particle and $\langle H \rangle$.

    First, we need to normalize the wave function to find the constant \( A \):
    $$        \int_0^L |\psi_n(x)|^2 \, dx = 1 $$
    This gives:
    $$        |A|^2 \int_0^L e^{2x/L} \, dx = 1 $$
    Evaluating the integral:
    $$        |A|^2 \left[ \frac{L}{2} e^{2x/L} \right]_0^L = 1 $$
    $$        |A|^2 \left( \frac{L}{2} (e^2 - 1) \right) = 1 $$
    Thus, we find:
    $$        |A|^2 = \frac{2}{L(e^2 - 1)} \implies A = \sqrt{\frac{2}{L(e^2 - 1)}} $$
    Now, we can calculate the expectation value of position \( \langle x \rangle \):
    $$        \langle x \rangle = \int_0^L \psi_n^*(x) \, x \, \psi_n(x) \, dx $$
    $$        = |A|^2 \int_0^L x e^{2x/L} \, dx $$
    Evaluating the integral using integration by parts:
    $$        \int x e^{2x/L} \, dx = \left(\frac{L}{2} x e^{2x/L} - \frac{L^2}{4} e^{2x/L}\right) + C $$
    Thus,
    $$        \langle x \rangle = |A|^2 \left[ \frac{L}{2} x e^{2x/L} - \frac{L^2}{4} e^{2x/L} \right]_0^L $$
    $$        = |A|^2 \left( \frac{L^2}{2} e^2 - \frac{L^2}{4} e^2 + \frac{L^2}{4} \right) $$
    $$        = |A|^2 \frac{L^2}{4} (e^2 + 1) $$
    Substituting \( |A|^2 \):
    $$        \langle x \rangle = \frac{2}{L(e^2 - 1)} \cdot \frac{L^2}{4} (e^2 + 1) = \frac{L(e^2 + 1)}{2(e^2 - 1)} $$
    Next, we calculate the expectation value of the Hamiltonian \( \langle H \rangle \):
    $$        \langle H \rangle = \int_0^L \psi_n^*(x) \, \hat{H} \, \psi_n(x) \, dx $$
    where the Hamiltonian operator \( \hat{H} \) for a particle in an infinite potential well is:
    $$        \hat{H} = -\frac{\hbar^2}{2m} \frac{d^2}{dx^2} $$
    Calculating the second derivative of \( \psi_n(x) \):
    $$        \frac{d^2 \psi_n(x)}{dx^2} = \frac{A}{L^2} e^{x/L} $$
    Thus,
    $$        \hat{H} \psi_n(x) = -\frac{\hbar^2}{2m} \cdot \frac{A}{L^2} e^{x/L} = -\frac{\hbar^2 A}{2m L^2} e^{x/L} $$
    Now, we can compute \( \langle H \rangle \):
    $$        \langle H \rangle = -\frac{\hbar^2 A^2}{2m L^2} \int_0^L e^{2x/L} \, dx $$
    Evaluating the integral:
    $$        \int_0^L e^{2x/L} \, dx = \left[ \frac{L}{2} e^{2x/L} \right]_0^L = \frac{L}{2} (e^2 - 1) $$
    Thus,
    $$        \langle H \rangle = -\frac{\hbar^2 A^2}{2m L^2} \cdot \frac{L}{2} (e^2 - 1) = -\frac{\hbar^2 A^2 (e^2 - 1)}{4m L} $$
    Substituting \( A^2 \):
    $$        \langle H \rangle = -\frac{\hbar^2}{4m L} \cdot \frac{2}{L(e^2 - 1)} (e^2 - 1) = -\frac{\hbar^2}{2m L^2} $$
    Therefore, the expectation values are:
    $$        \langle x \rangle = \frac{L(e^2 + 1)}{2(e^2 - 1)}, \quad \langle H \rangle = -\frac{\hbar^2}{2m L^2} $$
\end{example}
\end{document}
